---
title: "PTRS all-chromosome score concat — external CAMP cohorts (unified tissue / cell-type)"
output: html_document
---

Single parameterized RMD covering both external evaluation cohorts:

- `MODEL_VERSION = "tissue"` → 39 GTEx bulk-tissue PTRS on **CAMP+1KG**
  (fam: `CAMP_onek1k_chr1.fam`)
- `MODEL_VERSION = "ct"`     → 17 OneK1K cell-type PTRS on **CAMP+GTEx**
  (fam: `CAMP_GTEX_chr1.fam`)

This file concatenates per-chromosome PTRS scores across chromosomes 1-22 for
each feature, merges with the cohort's phenotype, and writes the per-feature
CSVs consumed by the unified notebooks (`05_*`):

- `combine/data/1/ptrs_results-concat_39_camp_1k1k/<tissue>_results.csv`
- `combine/data/1/ptrs_results-concat_17CT_camp_gtex/<cell_type>_results.csv`

It also writes the cohort phenotype to
`combine/data/updated_pheno_camp_<suffix>.txt`.

Flip `MODEL_VERSION` in the first chunk and re-knit to produce the other variant.

## 0. Config — flip `MODEL_VERSION` and re-run

```{r}
MODEL_VERSION <- "tissue"   # "tissue" or "ct"
stopifnot(MODEL_VERSION %in% c("tissue", "ct"))

CONFIG <- list(
  tissue = list(
    feature_label       = "tissue",
    feature_count_tag   = "39",
    feature_set_label   = "39 GTEx tissues",
    feature_source      = "${REPO_ROOT}/twas_path_table_39.csv",
    feature_source_type = "csv",   # "csv" → column 'tissue', or "listdir"
    base_dir            = "${REPO_ROOT}/outputs_39_focus_camp_1k1k",
    fam_file            = "${REPO_ROOT}/CAMP_onek1k_chr1.fam",
    pheno_out           = "${REPO_ROOT}/data/updated_pheno_camp_1k1k.txt",
    out_suffix          = "ptrs_results-concat_39_camp_1k1k",
    external_label      = "CAMP+1KG"
  ),
  ct = list(
    feature_label       = "cell_type",
    feature_count_tag   = "17CT",
    feature_set_label   = "17 OneK1K cell types",
    feature_source      = "${REPO_ROOT}/outputs_17CT_focus_camp_gtex/1",
    feature_source_type = "listdir",
    base_dir            = "${REPO_ROOT}/outputs_17CT_focus_camp_gtex",
    fam_file            = "${REPO_ROOT}/CAMP_GTEX_chr1.fam",
    pheno_out           = "${REPO_ROOT}/data/updated_pheno_camp_gtex.txt",
    out_suffix          = "ptrs_results-concat_17CT_camp_gtex",
    external_label      = "CAMP+GTEx"
  )
)
cfg <- CONFIG[[MODEL_VERSION]]
cat(sprintf("MODEL_VERSION = %s  (%s, external cohort = %s)\n",
            MODEL_VERSION, cfg$feature_set_label, cfg$external_label))
```

## 1. Libraries + helper

```{r}
library(data.table)
library(glmnet)
library(tidyverse)
library(ggplot2)
library(dplyr)
library(purrr)

process_data <- function(merged_data, vector_type) {
  data <- data.frame(
    prediction = merged_data[[vector_type]],
    asthma     = merged_data[["asthma"]]
  )
  group1 <- data$prediction[data$asthma == 1]
  group2 <- data$prediction[data$asthma == 0]
  t_test_result <- t.test(group1, group2)
  list(
    p_value = t_test_result$p.value,
    mean_x  = mean(group1, na.rm = TRUE),
    mean_y  = mean(group2, na.rm = TRUE)
  )
}
```

## 2. Build the feature list and per-feature input directories

```{r}
p_val <- "1"
chromosomes <- 1:22

features <- if (cfg$feature_source_type == "csv") {
  tbl <- fread(cfg$feature_source)
  sort(unique(tbl$tissue))
} else {
  sort(list.dirs(cfg$feature_source, recursive = FALSE, full.names = FALSE))
}
cat(sprintf("Loaded %d %s(s)\n", length(features), cfg$feature_label))

directories <- lapply(features, function(f) {
  path <- file.path(cfg$base_dir, p_val, f)
  if (!dir.exists(path)) dir.create(path, recursive = TRUE)
  path
})
names(directories) <- features
```

## 3. Load external cohort phenotype from fam (V6: 2=case, 1=control)

```{r}
fam <- read.table(cfg$fam_file, header = FALSE)
colnames(fam) <- c("V1", "V2", "V3", "V4", "V5", "V6")
fam$IID_base <- sub("_.*", "", fam$V2)
fam$asthma <- ifelse(fam$V6 == 2, 1, ifelse(fam$V6 == 1, 0, NA))

cat("asthma label distribution:\n")
print(table(fam$asthma, useNA = "ifany"))

write.table(
  fam,
  file = cfg$pheno_out,
  quote = FALSE, sep = " ", row.names = FALSE, col.names = TRUE
)
cat(sprintf("Wrote pheno -> %s\n", cfg$pheno_out))
```

## 4. Concat per-chromosome PTRS scores per feature, save merged CSVs

```{r}
pheno <- fread(cfg$pheno_out)

results_table <- data.frame(
  Tissue_Type = character(),
  Vector_Type = character(),
  Version     = character(),
  P_Value     = numeric(),
  Mean_X      = numeric(),
  Mean_Y      = numeric(),
  stringsAsFactors = FALSE
)

for (feat in names(directories)) {
  keep_vector    <- NULL
  sign_vector    <- NULL
  without_vector <- NULL
  keep_data      <- NULL  # holds last-read RDS for IID after chr loop

  for (chr in chromosomes) {
    directory    <- directories[[feat]]
    keep_file    <- paste0(directory, "/chr", chr, "_", feat, "_all_data_keep.rds")
    sign_file    <- paste0(directory, "/chr", chr, "_", feat, "_all_data_sign.rds")
    without_file <- paste0(directory, "/chr", chr, "_", feat, "_all_data.rds")

    if (file.exists(keep_file)) {
      keep_data <- readRDS(keep_file)
      if (is.null(keep_data)) { print(paste0(feat, chr)); next }
      keep_vector <- if (is.null(keep_vector)) keep_data$SCORE else keep_vector + keep_data$SCORE
    } else { print(paste0(feat, chr)); next }

    if (file.exists(sign_file)) {
      sign_data <- readRDS(sign_file)
      sign_vector <- if (is.null(sign_vector)) sign_data$SCORE else sign_vector + sign_data$SCORE
    } else { print(paste0(feat, chr)); next }

    if (file.exists(without_file)) {
      without_data <- readRDS(without_file)
      without_vector <- if (is.null(without_vector)) without_data$SCORE else without_vector + without_data$SCORE
    } else { print(paste0(feat, chr)); next }
  }

  if (is.null(keep_vector) || is.null(sign_vector) || is.null(without_vector)) {
    cat("Skipping", feat, "- no valid chromosome data\n"); next
  }

  sample_id <- keep_data$IID
  if (!(length(sample_id) == length(keep_vector) &&
        length(keep_vector) == length(sign_vector) &&
        length(sign_vector) == length(without_vector))) {
    cat("Skipping", feat, "- mismatched vector lengths\n"); next
  }

  combined_data <- data.frame(
    Sample_ID      = sample_id,
    Keep_Vector    = keep_vector,
    Sign_Vector    = sign_vector,
    Without_Vector = without_vector
  )
  merged_data <- merge(combined_data, pheno, by.x = "Sample_ID", by.y = "V2", all.x = TRUE)

  out_dir <- paste0(
    "${REPO_ROOT}/data/",
    p_val, "/", cfg$out_suffix
  )
  if (!dir.exists(out_dir)) dir.create(out_dir, recursive = TRUE)
  write.csv(merged_data, file.path(out_dir, paste0(feat, "_results.csv")), row.names = FALSE)

  for (vt in c("Keep", "Sign", "Without")) {
    vt_col <- paste0(vt, "_Vector")
    org <- process_data(merged_data, vt_col)
    results_table <- rbind(results_table,
      data.frame(Tissue_Type = feat, Vector_Type = vt, Version = "Original",
                 P_Value = org$p_value, Mean_X = org$mean_x, Mean_Y = org$mean_y))
  }
}

cat(sprintf("\nDone. External cohort = %s. Per-feature CSVs in:\n  combine/data/%s/%s/\n",
            cfg$external_label, p_val, cfg$out_suffix))
print(results_table)
```
