---
title: "PTRS all-chromosome score concat — GACRS (unified tissue / cell-type)"
output: html_document
---

Single parameterized RMD covering both:

- `MODEL_VERSION = "tissue"` → 39 GTEx bulk-tissue PTRS, scored on GACRS+1KG
- `MODEL_VERSION = "ct"`     → 17 OneK1K cell-type PTRS, scored on GACRS+1KG

For both versions this file:

1. Loads GACRS phenotype from the fam file + asthma case list
2. Splits GACRS into 80% train+test / 20% validation (seed = 42)
3. Concatenates per-chromosome PTRS scores (`Keep` / `Sign` / `Without` vectors)
   across chromosomes 1-22 for each feature
4. Writes per-feature merged CSVs that the unified notebooks (`05_*`) consume:
   - `combine/data/1/ptrs_results-concat_<count>_gacrs_train_test/<feature>_results.csv`
   - `combine/data/1/ptrs_results-concat_<count>_gacrs_val/<feature>_results.csv`
5. Runs the validation-set feature-selection analysis and writes
   `selected_<...>_all_case_gt_ctrl_keep.csv` — the focus list the notebooks read.

Flip `MODEL_VERSION` in the first chunk and re-knit to produce the other variant.

## 0. Config — flip `MODEL_VERSION` and re-run

```{r}
MODEL_VERSION <- "tissue"   # "tissue" or "ct"
stopifnot(MODEL_VERSION %in% c("tissue", "ct"))

CONFIG <- list(
  tissue = list(
    feature_label     = "tissue",
    feature_count_tag = "39",
    feature_set_label = "39 GTEx tissues",
    feature_source    = "${REPO_ROOT}/twas_path_table_39.csv",
    feature_source_type = "csv",   # "csv" → column 'tissue', or "listdir"
    base_dir          = "${REPO_ROOT}/outputs_39_focus_gacrs_1kg_new",
    out_suffix_prefix = "ptrs_results-concat_39_gacrs",
    summary_dir       = "${REPO_ROOT}/results_concat_39_new",
    summary_file_prefix = "ptrs_results_summary_gacrs",
    selected_prefix   = "selected_gacrs_val",
    selection_label   = "tissue",
    evaluation_file   = "evaluation_gacrs_val.csv",
    summary_long_file = "summary_tables_all_tissues_gacrs_val.csv",
    zoom_features     = c("Lung", "Whole_Blood")  # tissue-specific final density plots
  ),
  ct = list(
    feature_label     = "cell_type",
    feature_count_tag = "17CT",
    feature_set_label = "17 OneK1K cell types",
    feature_source    = "${REPO_ROOT}/outputs_17CT_focus_gacrs_1kg/1",
    feature_source_type = "listdir",
    base_dir          = "${REPO_ROOT}/outputs_17CT_focus_gacrs_1kg",
    out_suffix_prefix = "ptrs_results-concat_17CT_gacrs",
    summary_dir       = "${REPO_ROOT}/results_concat_17CT",
    summary_file_prefix = "ptrs_results_summary_17CT_gacrs",
    selected_prefix   = "selected_17CT_gacrs_val",
    selection_label   = "celltype",
    evaluation_file   = "evaluation_17CT_gacrs_val.csv",
    summary_long_file = "summary_tables_all_17CT_gacrs_val.csv",
    zoom_features     = character(0)  # skip per-feature zoom for CT
  )
)
cfg <- CONFIG[[MODEL_VERSION]]
cat(sprintf("MODEL_VERSION = %s  (%s)\n", MODEL_VERSION, cfg$feature_set_label))
```

## 1. Libraries + helpers

```{r}
library(data.table)
library(glmnet)
library(tidyverse)
library(ggplot2)
library(dplyr)
library(purrr)

# Welch t-test + means for one vector column (no residualization)
process_data <- function(merged_data, vector_type) {
  data <- data.frame(
    prediction = merged_data[[vector_type]],
    asthma     = merged_data[["asthma"]]
  )
  group1 <- data$prediction[data$asthma == 1]
  group2 <- data$prediction[data$asthma == 0]
  t_test_result <- t.test(group1, group2)
  list(
    p_value = t_test_result$p.value,
    mean_x  = mean(group1, na.rm = TRUE),
    mean_y  = mean(group2, na.rm = TRUE)
  )
}
```

## 2. Build the list of features (tissues or cell types) and per-feature output dirs

```{r}
p_val <- "1"

features <- if (cfg$feature_source_type == "csv") {
  tbl <- fread(cfg$feature_source)
  sort(unique(tbl$tissue))
} else {
  sort(list.dirs(cfg$feature_source, recursive = FALSE, full.names = FALSE))
}
cat(sprintf("Loaded %d %s(s)\n", length(features), cfg$feature_label))

# Per-feature input directory (where the per-chromosome RDS files live)
directories <- lapply(features, function(f) {
  path <- file.path(cfg$base_dir, p_val, f)
  if (!dir.exists(path)) dir.create(path, recursive = TRUE)
  path
})
names(directories) <- features
```

## 3. Load GACRS phenotype + label asthma cases

```{r}
gacrs_cases <- read.table(
  "${REPO_ROOT}/gacrs_cases.txt",
  header = FALSE, stringsAsFactors = FALSE
)
colnames(gacrs_cases) <- c("SUBJECT_ID", "NWD_ID")
head(gacrs_cases)

fam <- read.table(
  "${REPO_ROOT}/GACRS_1kg_chr1.fam",
  header = FALSE
)
fam$IID_base <- sub("_.*$", "", fam$V2)
# 1 → asthma case (subject in case list); 0 → not in case list
fam$asthma <- as.integer(fam$IID_base %in% gacrs_cases$SUBJECT_ID)
head(fam)
```

## 4. 80/20 train+test / validation split on GACRS (seed=42)

```{r}
fam_gacrs <- fam %>% filter(grepl("^CR", V2))

set.seed(42)
n_gacrs <- nrow(fam_gacrs)
val_idx <- sample(seq_len(n_gacrs), size = round(0.2 * n_gacrs), replace = FALSE)

fam_val        <- fam_gacrs[val_idx, ]
fam_train_test <- fam_gacrs[-val_idx, ]

cat("Validation:", nrow(fam_val), "samples\n")
cat("Train+Test:", nrow(fam_train_test), "samples\n")
cat("Val cases:",        sum(fam_val$asthma),
    "controls:",         sum(fam_val$asthma == 0), "\n")
cat("Train+Test cases:", sum(fam_train_test$asthma),
    "controls:",         sum(fam_train_test$asthma == 0), "\n")

write.table(fam_val,
            file = "${REPO_ROOT}/data/updated_pheno_gacrs_val.txt",
            quote = FALSE, sep = " ", row.names = FALSE, col.names = TRUE)
write.table(fam_train_test,
            file = "${REPO_ROOT}/data/updated_pheno_gacrs_train_test.txt",
            quote = FALSE, sep = " ", row.names = FALSE, col.names = TRUE)
```

## 5. Concat per-chromosome PTRS scores per feature, save merged CSVs

For each feature: sum `SCORE` across chromosomes 1-22 for each of the three
vectors (Keep / Sign / Without), merge with phenotype, write CSV used by the
unified notebooks.

```{r}
chromosomes <- 1:22

splits <- list(
  list(name = "val",
       pheno = fam_val,
       out_suffix = paste0(cfg$out_suffix_prefix, "_val"),
       summary_file = paste0(cfg$summary_file_prefix, "_val.csv")),
  list(name = "train_test",
       pheno = fam_train_test,
       out_suffix = paste0(cfg$out_suffix_prefix, "_train_test"),
       summary_file = paste0(cfg$summary_file_prefix, "_train_test.csv"))
)

for (sp in splits) {
  cat("\n=== Processing:", sp$name, "===\n")
  pheno <- sp$pheno

  results_table <- data.frame(
    Tissue_Type = character(),
    Vector_Type = character(),
    Version     = character(),
    P_Value     = numeric(),
    Mean_X      = numeric(),
    Mean_Y      = numeric(),
    stringsAsFactors = FALSE
  )

  for (feat in names(directories)) {
    keep_vector    <- NULL
    sign_vector    <- NULL
    without_vector <- NULL
    keep_data      <- NULL  # for sample_id after the chr loop

    for (chr in chromosomes) {
      directory     <- directories[[feat]]
      keep_file     <- paste0(directory, "/chr", chr, "_", feat, "_all_data_keep.rds")
      sign_file     <- paste0(directory, "/chr", chr, "_", feat, "_all_data_sign.rds")
      without_file  <- paste0(directory, "/chr", chr, "_", feat, "_all_data.rds")

      if (file.exists(keep_file)) {
        keep_data <- readRDS(keep_file)
        if (is.null(keep_data)) { print(paste0(feat, chr)); next }
        keep_vector <- if (is.null(keep_vector)) keep_data$SCORE else keep_vector + keep_data$SCORE
      } else { print(paste0(feat, chr)); next }

      if (file.exists(sign_file)) {
        sign_data <- readRDS(sign_file)
        sign_vector <- if (is.null(sign_vector)) sign_data$SCORE else sign_vector + sign_data$SCORE
      } else { print(paste0(feat, chr)); next }

      if (file.exists(without_file)) {
        without_data <- readRDS(without_file)
        without_vector <- if (is.null(without_vector)) without_data$SCORE else without_vector + without_data$SCORE
      } else { print(paste0(feat, chr)); next }
    }

    if (is.null(keep_vector) || is.null(sign_vector) || is.null(without_vector)) {
      cat("Skipping", feat, "- no valid chromosome data\n"); next
    }

    sample_id <- keep_data$IID
    if (!(length(sample_id) == length(keep_vector) &&
          length(keep_vector) == length(sign_vector) &&
          length(sign_vector) == length(without_vector))) {
      cat("Skipping", feat, "- mismatched vector lengths\n"); next
    }

    combined_data <- data.frame(
      Sample_ID      = sample_id,
      Keep_Vector    = keep_vector,
      Sign_Vector    = sign_vector,
      Without_Vector = without_vector
    )
    merged_data <- merge(combined_data, pheno, by.x = "Sample_ID", by.y = "V2", all = FALSE)

    if (nrow(merged_data) == 0 || length(unique(merged_data$asthma)) < 2) {
      cat("Skipping", feat, "for", sp$name, "- insufficient data\n"); next
    }

    out_dir <- paste0(
      "${REPO_ROOT}/data/",
      p_val, "/", sp$out_suffix
    )
    if (!dir.exists(out_dir)) dir.create(out_dir, recursive = TRUE)
    write.csv(merged_data, file.path(out_dir, paste0(feat, "_results.csv")), row.names = FALSE)

    for (vt in c("Keep", "Sign", "Without")) {
      vt_col <- paste0(vt, "_Vector")
      org <- process_data(merged_data, vt_col)
      results_table <- rbind(results_table,
        data.frame(Tissue_Type = feat, Vector_Type = vt, Version = "Original",
                   P_Value = org$p_value, Mean_X = org$mean_x, Mean_Y = org$mean_y))
    }
  }

  cat("\n", sp$name, "results:\n"); print(results_table)

  results_original_sorted <- results_table %>%
    filter(Version == "Original") %>%
    arrange(P_Value)

  summary_dir <- file.path(cfg$summary_dir, p_val)
  if (!dir.exists(summary_dir)) dir.create(summary_dir, recursive = TRUE)
  write.csv(results_original_sorted,
            file.path(summary_dir, sp$summary_file), row.names = FALSE)
}
```

## 6. Validation-set feature selection — derived files used by the notebooks

The per-split summary CSV produced above is the input. From the validation
summary we derive three selection files (all written under `cfg$summary_dir`):

- `<selected_prefix>_by_best_tissue_keep.csv`      — features with Keep-vector case>control and P<0.5
- `<selected_prefix>_top10_keep.csv`               — top-10 of the above by P
- `<selected_prefix>_all_case_gt_ctrl_keep.csv`    — **consumed by the notebooks** (no P filter)

```{r}
val_summary_path <- file.path(
  cfg$summary_dir, p_val,
  paste0(cfg$summary_file_prefix, "_val.csv")
)
results_original_sorted <- fread(val_summary_path) %>% arrange(P_Value)
head(results_original_sorted)

BASE_DIR <- cfg$summary_dir
p_vals   <- c("1")

versions <- tibble(
  PTRS_Version = c("GACRS Val"),
  file         = c(paste0(cfg$summary_file_prefix, "_val.csv"))
)
```

```{r}
evaluation_df <- versions %>%
  pmap_dfr(function(PTRS_Version, file) {
    bind_rows(lapply(p_vals, function(p_val) {
      path <- file.path(BASE_DIR, p_val, file)
      df <- fread(path)
      df %>%
        mutate(
          Better      = Mean_X > Mean_Y,
          Significant = P_Value < 0.5,
          p_val       = p_val,
          PTRS_Version = PTRS_Version
        ) %>%
        group_by(PTRS_Version, p_val, Vector_Type) %>%
        summarise(
          N_MeanX_gt_MeanY     = sum(Better),
          N_MeanX_gt_MeanY_sig = sum(Better & Significant),
          .groups = "drop"
        )
    }))
  })

evaluation_df

fwrite(
  evaluation_df,
  file = file.path(BASE_DIR, cfg$evaluation_file)
)
```

```{r}
summary_tables <- versions %>%
  pmap(function(PTRS_Version, file) {
    tables <- lapply(p_vals, function(p_val) {
      path <- file.path(BASE_DIR, p_val, file)
      df <- fread(path)
      df %>%
        filter(Mean_X > Mean_Y, P_Value < 0.5) %>%
        arrange(P_Value) %>%
        mutate(p_val = p_val, Version = PTRS_Version)
    })
    names(tables) <- p_vals
    tables
  })
names(summary_tables) <- versions$PTRS_Version

summary_long <- bind_rows(lapply(names(summary_tables), function(v) bind_rows(summary_tables[[v]])))

fwrite(
  summary_long,
  file = file.path(BASE_DIR, cfg$summary_long_file)
)
```

```{r}
top10_by_pval <- versions %>%
  pmap(function(PTRS_Version, file) {
    tables <- lapply(p_vals, function(p_val) {
      path <- file.path(BASE_DIR, p_val, file)
      df <- fread(path)
      df %>%
        filter(Mean_X > Mean_Y, P_Value < 0.5) %>%
        group_by(Tissue_Type) %>%
        slice_min(order_by = P_Value, n = 1, with_ties = FALSE) %>%
        ungroup() %>%
        arrange(P_Value) %>%
        slice_head(n = 10) %>%
        mutate(p_val = p_val, Version = PTRS_Version)
    })
    names(tables) <- p_vals
    tables
  })
names(top10_by_pval) <- versions$PTRS_Version
```

```{r}
evaluation_df <- evaluation_df %>%
  mutate(pval_vector = paste(p_val, Vector_Type, sep = " | "))

ggplot(
  evaluation_df,
  aes(x = pval_vector, y = N_MeanX_gt_MeanY_sig, fill = PTRS_Version)
) +
  geom_col(position = position_dodge(width = 0.75), width = 0.65) +
  labs(
    x     = "p-value threshold | Vector type",
    y     = sprintf("Number of %ss (Mean_X > Mean_Y & P < 0.5)", cfg$feature_label),
    title = sprintf("Significant %ss (GACRS Validation Set, %s)",
                    cfg$feature_label, cfg$feature_set_label)
  ) +
  theme_bw() +
  theme(axis.text.x = element_text(angle = 45, hjust = 1),
        legend.position = "top",
        panel.grid.major.x = element_blank())
```

```{r}
p_val_use <- "1"
df_gacrs <- top10_by_pval[["GACRS Val"]][[p_val_use]] %>%
  select(Tissue_Type, P_Value) %>%
  rename(P_GACRS = P_Value)

ggplot(
  df_gacrs,
  aes(x = -log10(P_GACRS), y = reorder(Tissue_Type, -log10(P_GACRS)))
) +
  geom_point(size = 3, color = "#08519C") +
  geom_vline(xintercept = -log10(0.05), linetype = "dashed", alpha = 0.4) +
  labs(
    title = sprintf("GACRS Val (%s) — %s", p_val_use, cfg$feature_set_label),
    x     = expression(-log[10](P)),
    y     = stringr::str_to_title(cfg$feature_label)
  ) +
  theme_bw()
```

```{r}
# Write the three selection CSVs
selected <- results_original_sorted %>%
  filter(Vector_Type == "Keep", Mean_X > Mean_Y, P_Value < 0.5) %>%
  select(Tissue_Type, Vector_Type, Version, P_Value, Mean_X, Mean_Y)
write.csv(
  selected,
  file.path(cfg$summary_dir, paste0(cfg$selected_prefix, "_by_best_", cfg$selection_label, "_keep.csv")),
  row.names = FALSE
)

top10 <- selected %>% arrange(P_Value) %>% head(10)
write.csv(
  top10,
  file.path(cfg$summary_dir, paste0(cfg$selected_prefix, "_top10_keep.csv")),
  row.names = FALSE
)

# all_case_gt_ctrl: NO P-value filter — this is the file the notebooks read
all_case_gt_ctrl <- results_original_sorted %>%
  filter(Vector_Type == "Keep", Mean_X > Mean_Y) %>%
  select(Tissue_Type, Vector_Type, Version, P_Value, Mean_X, Mean_Y)
write.csv(
  all_case_gt_ctrl,
  file.path(cfg$summary_dir, paste0(cfg$selected_prefix, "_all_case_gt_ctrl_keep.csv")),
  row.names = FALSE
)

cat(sprintf("Total %ss with case > control: %d\n",
            cfg$feature_label, nrow(all_case_gt_ctrl)))
all_case_gt_ctrl
```

## 7. (Optional) Per-feature zoom-in density plots

Only runs for `MODEL_VERSION = "tissue"` (uses `Lung` and `Whole_Blood`). Set
`cfg$zoom_features` in the config to enable for other variants.

```{r}
if (length(cfg$zoom_features) >= 2) {
  library(patchwork)

  zoom_dir <- paste0(
    "${REPO_ROOT}/data/",
    p_val, "/", cfg$out_suffix_prefix, "_val"
  )

  f1 <- cfg$zoom_features[1]
  f2 <- cfg$zoom_features[2]
  df1 <- read.csv(file.path(zoom_dir, paste0(f1, "_results.csv")))
  df2 <- read.csv(file.path(zoom_dir, paste0(f2, "_results.csv")))
  df1$asthma <- factor(df1$asthma, levels = c(0, 1), labels = c("Control", "Case"))
  df2$asthma <- factor(df2$asthma, levels = c(0, 1), labels = c("Control", "Case"))

  p1 <- ggplot(df1, aes(x = Keep_Vector, fill = asthma)) + geom_density(alpha = 0.5) +
    labs(title = sprintf("%s - Keep PTRS", f1), x = "PTRS Score", y = "Density") + theme_bw()
  p2 <- ggplot(df1, aes(x = Sign_Vector, fill = asthma)) + geom_density(alpha = 0.5) +
    labs(title = sprintf("%s - Sign PTRS", f1), x = "PTRS Score", y = "Density") + theme_bw()
  p3 <- ggplot(df2, aes(x = Keep_Vector, fill = asthma)) + geom_density(alpha = 0.5) +
    labs(title = sprintf("%s - Keep PTRS", f2), x = "PTRS Score", y = "Density") + theme_bw()
  p4 <- ggplot(df2, aes(x = Sign_Vector, fill = asthma)) + geom_density(alpha = 0.5) +
    labs(title = sprintf("%s - Sign PTRS", f2), x = "PTRS Score", y = "Density") + theme_bw()

  (p1 | p2) / (p3 | p4)
} else {
  cat("Skipping zoom-in plots (cfg$zoom_features empty for MODEL_VERSION =", MODEL_VERSION, ")\n")
}
```
